Characterization of the microbial and fungal communities in ancient pottery from Neolithic southwestern China

### 2024-01-31 analysis/paper/paper.qmd
work on method

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China. As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting. The post-excavation analysis of Guijiabao is still in progress. This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology. Through a comparative analysis of different parts of ceramic samples and soil samples, we aim to answer the research question: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? Our hypothesis is that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective sheds light on food practices from a microbial perspective that has not been fully discussed in archaeology.

@@

To minimize potential sources of contamination from the environments, we cleaned all lab benches and pipettors using 99% ethanol and UV light before sample preparation and DNA extraction. These procedures were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan. Because contemporary contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions and reagents, and consumables. A total of 12 powdered pottery samples (interior and exterior parts), five soil samples from Guijiabao and Gaoshan, and three negative control samples were tested in this study.

## sample preparation

Each pottery sample was divided into two distinct portions: the interior, representing the middle part of the potsherd matrix, and the exterior surface exposed to the environment. This division aimed to facilitate a comparative analysis of the microbial communities in residues absorbed into the potsherds versus those on the surface. The collection of approximately 1g of the interior potsherd was executed using a sterile drill operated at a low speed to prevent heat generation. The powdered sample was initially gathered on a pre-combustion foil and subsequently transferred into pre-sterilized tubes. A parallel sampling procedure was conducted for the surface of the potsherds, with the drill being cleaned between each iteration to minimize cross-contamination. For the soil samples, around 1g was collected from the tube containing the soil samples using sterile spatulas.

## DNA extraction and amplification

Total DNA was extracted for each sample using the DNeasy Power Soil Pro Kit (QIAGEN). All DNAs were quantified using Qubit Fluorometer.

Focus on bacterial, archaeal and fungal communities
Primer selection for target regions (100-300 bp): 
16S rRNA gene (V1, V4, V6) 
18S rRNA gene (ITS1)
Amplification with PCR
Gel running and collecting the bands of regions
Illumina amplicon sequencing

Amplicons of V1, V4, and V6 of 16S and ITS1 of ITS1.

>>>>

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China. As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting. The post-excavation analysis of Guijiabao is still in progress. This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology. We aim to explore the differences in overall microbial communities, including bacterial, archaeal and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples. Our main research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? We hypothesized that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective could shed light on food practices from a microbial perspective that has not been fully discussed in archaeology.

@@

To minimize potential sources of contamination from the environments, we cleaned all lab benches, equipment, and pipettors with 3% bleach, followed by 99% ethanol and UV light. The sample preparation and DNA extraction were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan. Because modern contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions, reagents, and consumables. A total of 12 powdered pottery samples (interior and exterior parts), 5 soil samples from Guijiabao and Gaoshan, and 3 negative control samples were analyzed in this study.

## Sample preparation and DNA extracation

Each pottery sample was divided into two portions: the interior, representing the middle part of the potsherd matrix, and the exterior surface exposed to the environment. This division aimed to facilitate a comparative analysis of the microbial communities in residues absorbed into the potsherds versus those on the surface. Approximately 1.5 g of each sample was collected using a sterile drill operated at a low speed to prevent heat generation. The powdered sample was initially gathered on pre-combustion foil and immediately transferred to a sterile Eppendorf tube. All the tools were cleaned between each iteration to minimize cross-contamination. For the soil samples, around 1g was collected using sterile spatulas and transferred to a sterile Eppendorf tube.

Total DNA from each sample was extracted in a portion of 300 mg using the DNeasy PowerSoil Pro kit (Qiagen, Germany), and the DNA extracts from the same sample were pooled. The concentration of DNA was measured using a Qubit 2.0 fluorometer and standardized to 1 ng μL^−1^.

## PCR analyses

Given the considerable fragmentation of DNA in archaeological contexts, we focused on short regions ranging from 100 to 300 bp, including V1, V4, and V6 of 16S rRNA and ITS1 of 18S rRNA. For bacterial communities, the V1 region was amplified using the primers 8F (5'-AGAGTTTGATYMTGGCTCAG-3') and 120R (5'-TTACTCACCCGTICGCCRCT-3') [@]. For archaeal communities, V4 region was amplified using arch519F (5′-CAGCMGCCGCGGTAA-3′) and arch806R (5′-GGACTACVSGGGTATCTAAT-3′), and V6 region was amplified using 926F (5’-AAACTYAAAKGAATTGACGGG-3’) and 1046R (5’-CGACARCCATGCASCACCT-3’) [@Siles2018]. For fungal communities, ITS1 region of 18S rRNA was amplified using the primers ITS1F (5’-CTTGGTCATTTAGAGGAAGTAA-3’) and ITS2 (5′-GCTGCGTTCTTCATCGATGC-3′) [@Siles2018]. Barcoded forward adapters were added to the PCR primers to facilitate sample pooling for library preparations. The DNA amplification for bacteria was conducted using the Taq DNA polymerase Master Mix RED (Ampliqon) under the following program: initial denaturation at 95 °C for 1 min, followed by 35 cycles of denaturation at 95 °C for 30 s, primer annealing at 50 °C for 30 s, extension at 72 °C for 30 s, and final elongation at 72 °C for 5 min. For fungi, the thermal conditions are the same as for bacteria, with the exception of 40 cycles and annealing temperature at 57 °C for 40 s.

The DNA fragments in PCR products were purified with the GenepHlow gel extraction kit  (Geneaid Biotech Ltd.) after conducting 2 % agarose gel electrophoresis. The amplified products were then pooled and used to build a DNA library using the Illumina TruSeq DNA library preparation protocol.
